---
title: "Data Wrangling"
author: "From Raw Data to Regional Insights: Intermediate R Tools for Workforce & Economic Development"
date: "Nov 19-20, 2025"
output:
  html_document:
    css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
    number_sections: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


The tidyverse provides a consistent grammar for cleaning, transforming, reshaping, and joining data. This section uses a **custom workforce dataset** to illustrate practical transformations that analysts perform in economic and labor market research.


```{r}
library(tidyverse)

workforce_example <- tibble(
  county = c("Mercer", "Middlesex", "Morris", "Union", "Somerset"),
  jobs_2023 = c(185200, 401500, 302000, 292100, 178300),
  jobs_2024 = c(188400, 407200, 309500, 295300, 181900),
  wage = c(34.2, 36.8, 38.1, 35.4, 37.2),
  unemp_rate = c(4.1, 3.8, 3.3, 4.5, 3.7),
  ui_claims = c(1540, 2100, 1700, 2500, 1200),
  openings = c(8200, 15100, 9200, 10300, 6100)
)

workforce_example
```

---

# 1. Pipes: The Core of Tidyverse Workflow

Pipes streamline multi-step transformations.

```{r}
workforce_example |>
  filter(unemp_rate > 4) |>
  mutate(openings_ratio = openings / jobs_2024) |>
  summarize(avg_ratio = mean(openings_ratio))
```

✔ Pipes will be used throughout this workshop in FRED, BLS, ACS, PUMS, and mapping workflows.

---

### 📘 Exercises
1. Filter for counties with wage > 36.  
2. Create `jobs_growth = jobs_2024 - jobs_2023`.  
3. Compute average unemployment rate across counties.

---

# 2. Selecting Columns with `select()`

```{r}
workforce_example |> 
  select(county, wage, unemp_rate)
```

Select columns by pattern:

```{r, eval=FALSE}
select(workforce_example, starts_with("jobs"))
select(workforce_example, ends_with("rate"))
select(workforce_example, -ui_claims)
```

---

### 📘 Exercises
1. Select all job-related columns.  
2. Drop all columns except county and wage.

---

# 3. Filtering Rows with `filter()`

```{r}
workforce_example |>
  filter(jobs_2024 > 300000)
```

Multiple conditions:

```{r, eval=FALSE}
filter(workforce_example,
       wage > 36,
       unemp_rate < 4)
```

---

### 📘 Exercises
1. Filter counties with both high wages (> 37) **and** low unemployment (< 4).  
2. Filter counties with UI claims above 2000.  

---

# 4. Transforming Data with `mutate()`

```{r}
workforce_example |>
  mutate(
    jobs_growth = jobs_2024 - jobs_2023,
    wage_annual = wage * 2080
  )
```

Categorical variables:

```{r}
workforce_example |>
  mutate(
    category = case_when(
      unemp_rate < 3.5 ~ "Low unemployment",
      unemp_rate < 4.0 ~ "Moderate unemployment",
      TRUE ~ "High unemployment"
    )
  )
```

---

### 📘 Exercises
1. Create a `tightness = openings / ui_claims` variable.  
2. Create a `high_wage` flag where wage > 37.

---

# 5. Summaries with `group_by()` and `summarize()`

```{r}
workforce_example |>
  summarize(
    avg_wage = mean(wage),
    total_jobs_2024 = sum(jobs_2024)
  )
```

Group by custom classes:

```{r}
workforce_example |>
  mutate(size = if_else(jobs_2024 > 250000, "Large", "Small")) |>
  group_by(size) |>
  summarize(
    average_wage = mean(wage),
    avg_unemp = mean(unemp_rate)
  )
```

---

### 📘 Exercises
1. Group counties into high/low UI claims and summarize unemployment rate.  
2. Compute job growth by category from Section 4.

---

# 6. Sorting and Ordering

```{r}
workforce_example |> arrange(wage)
```

Descending:

```{r}
workforce_example |> arrange(desc(jobs_2024))
```

---

### 📘 Exercises
1. Sort counties by unemployment rate (highest to lowest).  
2. Sort by wage descending, then jobs ascending.

---

# 7. Reshaping Data (pivot_longer / pivot_wider)

## Wide → Long

```{r}
jobs_long <- workforce_example |>
  pivot_longer(
    cols = starts_with("jobs_"),
    names_to = "year",
    values_to = "jobs"
  )

jobs_long
```

## Long → Wide

```{r}
jobs_long |>
  pivot_wider(
    names_from = year,
    values_from = jobs
  )
```

---

### 📘 Exercises
1. Convert the wage and unemployment columns to long format.  
2. Create a clean year variable from `"jobs_2024"`.

---

# 8. Joining Data

Create a second dataset:

```{r}
metro_region <- tibble(
  county = c("Mercer","Union","Somerset"),
  metro = c("Trenton", "New York", "New York")
)
```

Left join:

```{r}
left_join(workforce_example, metro_region, by="county")
```

Inner join:

```{r}
inner_join(workforce_example, metro_region, by="county")
```

---

### 📘 Exercises
1. Add a new county to `metro_region` and observe `full_join()`.  
2. Explain which join preserves all counties from `workforce_example`.

---

# 9. Handling Missing Data

```{r}
df_na <- tibble(
  county = c("Mercer","Union","Somerset"),
  x = c(1, NA, 3),
  y = c(NA, 2, 5)
)

df_na |> drop_na()
df_na |> replace_na(list(x = 0, y = 0))
```

---

### 📘 Exercises
1. Count missing values with `summarize(across())`.  
2. Impute missing x values with the mean of x.

---

# 10. Strings and Dates (Common for BLS/ACS Workflows)

```{r}
library(stringr)
str_to_upper("mercer")
str_remove("jobs_2024", "jobs_")
```

```{r}
library(lubridate)
ymd("2024-03-01")
```

---

# Wrap-Up

This session introduced the core data wrangling workflow that underpins all analysis in this workshop:

- Pipes for clear multi-step transformations  
- Selecting, filtering, reshaping, and transforming data  
- Summarizing and grouping  
- Joining datasets  
- Handling missing values  
- Working with dates and strings  

These techniques prepare you for more advanced applications using FRED, BLS, ACS, PUMS, and geospatial data.

